A Dissection of Foregrounds in Epoch of Reionization Observations from the Murchison Widefield Array

### 2014-11-14 ms.tex
Renamed the tex, ps, and pdf files to ms.tex, ms.ps and ms.pdf respectively

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact emission dominates the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space. Based on these results, we propose that selective down--weighting of baselines based on length, direction, and time will mitigate foreground contamination substantially (up to a factor $\sim 10$ with negligible loss of sensitivity) in redshifted 21~cm power spectrum analysis.

@@

\keywords{large-scale structure of Universe --- methods: statistical --- radio continuum: galaxies --- radio lines: general --- reionization --- techniques: interferometric}

@@

After matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds that most severely obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a significant role hitherto unpredicted. % We find that the foregrounds most likely to obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon. We find that diffuse Galactic emission plays a larger role then previously thought. 
We quantify this contamination by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

% Hence, we have identified a region of high significance (shown in medium shade) that forms a subset of the {\it EoR window} bounded by $|\tau|\le 0.78\,\mu$s where contamination from foregrounds is expected to be less and yet the EoR H{\sc i} power spectrum signal is reasonably strong. Hereafter, we refer to this region as the {\it EoR window} in the rest of this paper.

@@

% The terms antenna spacing and baseline are synonymous in radio interferometry. The latter term is also used broadly to refer to the antenna pair system in a radio interferometer array. We use these terms interchangeably in this paper.

@@

% The bandpass properties used in this paper are illustrated in Figure~\ref{fig:bandpass}. The MWA bandpass of width 30.72~MHz is divided coarsely into 24 1.28~MHz chunks (dashed lines connecting solid circles) with each chunk weighted by a digital filter. The coarse channel shape is obtained using a 8--tap polyphase filter bank (PFB) and a Kaiser window with parameter $\beta=5$. The bandpass correction ({\it plus} symbols connected by dotted lines) corrects for the shape of the coarse channels. After this correction, the channels at the edges of these chunks have been flagged (solid lines) because they are known to possibly be contaminated by aliasing at a low level. Any spectral flags, or other known weights constitutes $W_f(f)$ in equation~\ref{eqn:obsvis}.

% \begin{figure}[htb]
% \centering
% \includegraphics[width=\linewidth]{figures/v1_0/bandpass_properties.eps}
% \caption{Bandpass properties used in modeling and analysis of delay spectrum of MWA visibilities. The filled circles joined by dotted lines show the 8--tap PFB bandpass shapes obtained using a Kaiser window with parameter $\beta=5$. The bandpass correction factors applied are shown ({\it plus} symbols connected by dashed lines). The solid line shows the flagged (gaps) and gain--corrected bandpass shape. These shapes are repeated every 1.28~MHz for the entire bandwidth of 30.72~MHz centered around 185~MHz. \label{fig:bandpass}}
% \end{figure}

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=95$~K per polarization for all frequency channels for all antenna pairs throughout the course of the observation. Estimation of $T_\textrm{sys}$ from MWA data is a subject of active investigation. In our work, we have chosen a $T_\textrm{sys}$ that matches the thermal noise observed in the data. % We take into account the increase in thermal uncertainty in frequency channels that results due to aforementioned bandpass correction. ({\bf Since the bandpass correction is digital, shouldn't the noise stay flat after bandpass correction? -- JDB: yes})

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0h, Dec = $-30$\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of the dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg~ grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations shown here were undertaken on Aug~23, 2013. % Data set from this day has been chosen as a reference for comparison within the MWA EoR collaboration because this is one of the days featuring a low number of known system errors from among a long stretch of stable observing.  
We have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of a nominal observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at $-1.92$~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively. %These pointings (phased array power--patterns) are centered at RA = 4\fdg 387, Dec = -29\fdg 86 and RA = -28\fdg 8, Dec = -26\fdg 701, whereas the visibilities themselves are phased to zenith corresponding to RA = -28\fdg 8, Dec = -26\fdg 701, and RA = 1\fdg 35, Dec = -26\fdg 701 respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST). The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA). Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg~ or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0h, Dec = $-30$\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
\end{figure}

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off10,off12}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey \citep[MWACS;][]{hur14}. % MWACS was compiled using observations taken with subsets of the array during the final buildout and commissioning;
It has a flux density limit of 25~mJy and a declination range of -12\arcdeg~ to -40\arcdeg~ evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on forward modeling software by \citet{sul12} and the calibration method described by \citet{sal14} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are fairly low signal--to--noise so we iteratively average along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variations in cable delay. Finally, we fit for an additional phase known to be caused by small reflections in subset of cables.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two observations of the EoR field (RA = 0h, Dec = $-30$\arcdeg) at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. For reference, the $k$--space equivalent of $\tau$ and $|\vec{b}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The logarithmic color scale (shown at the right) is common to both panels. \label{fig:fhd_data}}
\end{figure}

% Despite delay deconvolution, we still see leakage beyond the cleaned regions indicating the inability of the deconvolution algorithm to fit the measurements perfectly, dominant at short antenna spacings where the number of delay modes within the horizon is lesser. This is because the maximum delay envelope (boundary of the wedge) consists of mixed emission from large portions of the sky into a narrow range of delays and the deconvolution algorithm does not have sufficient support in delay space to fit the delay spectrum. However, we wish to emphasize that the purpose of deconvolution is to rid the delay spectra of instrumental artifacts to the best extent possible in order to see the underlying signatures of foreground emission. It is not our intention in this paper to use the deconvolution as a foreground removal tool.

@@

The degree to which these objects leak power into spectral line of sight modes in the {\it EoR window} will be shown to depend primarily on the brightness of foregrounds near the boundaries of the {\it foreground wedge}. Thus, it is important to consider an all--sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting only to the primary field of view (a point also supported by Pober~et~al.~(2014), in prep.).

% JDB: I don't think we need this here. %% \citet{bea13} estimated the sensitivity of the MWA to EoR H{\sc i} power spectrum detection taking into account thermal noise effects in the {\it EoR window}. \citet{thy13} estimated the sensitivity taking into account the spillover from foreground contamination from unsubtracted extragalactic compact objects in the {\it EoR window} besides thermal noise. In this paper, we include the diffuse Galactic emission for enhancing our understanding of foreground signatures in the power spectrum.

@@

% \item {\it Instrumental Passband}: The instrumental passband characteristics have a direct effect on the delay transform owing to the Fourier relationship between the two. Due to excision of aliased edge channels repeated in each coarse channel of our passband, we observed repeated patterns of the {\it foreground wedge} in the delay spectra. As already mentioned earlier, we have used a deconvolution algorithm ({\it CLEAN}) to rid the delay spectra of band shape effects. Since this paper focuses on studying the effects of foreground on the delay spectra, we do not pursue the characterization of artifacts from the band shape and imperfections in {\it CLEAN} deconvolution.

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\vec{b}|\le$~200~m). This is contrary to expectations because these longer baselines are not expected to be sensitive to diffuse emission (on scales $\gtrsim$~0\fdg 85) of our input foreground model. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.   %% This contributes to a {\it pitchfork}--shaped feature which will be discussed later in \S\ref{sec:composite} in more detail.

@@

% While the {\it foreground wedge} is a region of high foreground contamination, of particular interest to the EoR community is foreground contamination near horizon delay limits. These regions are believed to be responsible for significant spillover into the relatively clean {\it EoR window} due to spectral response of the instrument and foregrounds.

@@

In principle, instead of discarding selected antenna spacings altogether, we could down--weight them based on an optimal scheme. This technique provides a very simple and yet effective tool in adding a layer of control to mitigate effects of foreground contamination in EoR data analysis. % The success behind clear identification of severely foreground--contaminated antenna pairs based on delay spectrum technique depends on the localization of the source of foreground contamination and its strength.

@@

In conclusion, we find that inclusion of emission models, both diffuse and compact, all the way to the horizon is essential to explaining the observed power spectrum. We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination by an order of magnitude in EoR data analysis by discarding or down--weighting antenna pairs most affected by foreground contamination as a function of antenna spacing vector and time of observation, with negligible loss of sensitivity.

>>>>

Detection of 21~cm emission of neutral hydrogen from the epoch of reionization, at redshifts $z>6$, is limited primarily by foreground emission.  We investigate the chromatic signatures of an all--sky foreground model imprinted by an instrumental transfer function.  Using the delay spectrum technique, we demonstrate that the foreground signatures that have the largest impact on the H{\sc i} signal arise from power received far away from the primary field of view. Comparing data from recent Murchison Widefield Array observations with simulations separated into components based on type of emission, we identify diffuse emission near the horizon as a significant contributing factor, even on wide antenna spacings. For signals entering through the primary field of view, compact emission dominates the foreground contamination. These two mechanisms imprint a characteristic {\it pitchfork} signature on the ``foreground wedge'' in Fourier space. Based on these results, we propose that selective down--weighting of data based on antenna spacing vector and time will mitigate foreground contamination substantially (up to a factor $\sim 10$ with negligible loss of sensitivity) in redshifted 21~cm power spectrum analysis.

@@

\keywords{cosmology: observations --- large-scale structure of universe --- methods: statistical --- radio continuum: galaxies --- reionization --- techniques: interferometric}

@@

After matching real observations from the Murchison Widefield Array with instrumental simulations of the entire known sky, we examine the signature of the power spectrum of each baseline. We report two important findings: the foregrounds that most severely obfuscate the redshifted 21~cm power spectrum are not caused by emission in the central field of view, but rather by bright objects from near the horizon; and, diffuse Galactic emission plays a significant role hitherto unpredicted. We quantify this contamination by separating the simulations into components based on type of foreground emission. We then arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to down--weight adversely contaminated baselines.

@@

@@

@@

Thermal noise in simulated visibilities is estimated assuming a system temperature of $T_\textrm{sys}=95$~K per polarization for all frequency channels for all antenna pairs throughout the course of the observation. Estimation of $T_\textrm{sys}$ from MWA data is a subject of active investigation. In our work, we have chosen a $T_\textrm{sys}$ that matches the thermal noise observed in the data.

@@

The MWA targets two primary low-foreground fields for reionization observations. Here, we focus on the field at RA = 0$^\textrm{h}$, Dec = $-30$\arcdeg. The MWA tracks a patch of sky through antenna beams formed and steered electronically by controlling delay settings of the dipoles in an MWA tile. The pointing system is capable of steering to points on a regular $\sim$7\arcdeg~ grid. During observations we allow the sky to drift across the nearest available pointing, shifting between grid points as necessary (once in $\sim 30$~minutes). This process is repeated throughout the course of the observation $\approx 4.86$~hours. Figure~\ref{fig:pointings} illustrates how the apparent pointing oscillates around the desired pointing direction.

The observations shown here were undertaken on 2013~August~23. % Data set from this day has been chosen as a reference for comparison within the MWA EoR collaboration because this is one of the days featuring a low number of known system errors from among a long stretch of stable observing.  
We have chosen two, 112 second long sections from this night for detailed study. These were chosen to provide a selection of possible foreground and instrumental conditions. As an example of a nominal observing setup we choose a zenith pointing; as an example of poor foreground conditions, we choose a pointing when the field is $\sim 2$~hours from zenith. This pointing has a significantly higher secondary lobe structure and is observed when the bright galactic center is well above the horizon.

The {\it local sidereal time} (LST) at these two snapshots are shown in Figure~\ref{fig:pointings} as vertical lines at $-1.92$~hours (i.e., 22.08~hours) and 0.09~hours which are hereafter denoted as {\it off--zenith} and {\it zenith} pointings, respectively.

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/custom_pointings.eps}
\caption{MWA tile beam pointing directions during the course of the observation. The $x$--axis refers to the {\it local sidereal time} (LST). The axis on the left refers to longitudes, namely, Right Ascension (RA) and Hour Angle (HA). Negative values of RA, HA, and LST are to be interpreted as having been wrapped around by 360\arcdeg~ or 24 hours. The axis on the right refers to the declination of the pointing direction. The RA, HA, and declination are plotted with solid, dashed and dotted lines respectively. The sky is allowed to drift for $\sim 30$~mins before the beamformer delay settings change in a discrete step to center the beam around RA = 0$^\textrm{h}$, Dec = $-30$\arcdeg. Dot--dashed vertical lines show the LST of {\it off-zenith} and {\it zenith} pointings used in our study. \label{fig:pointings}}
\end{figure}

@@

The data are flagged\footnote{\url{http://sourceforge.net/p/aoflagger}} for interference \citep{off10,off12}, removing 3\% of the data and averaged in time  and frequency from the raw 0.5~s, 40~kHz to 2s, 80~kHz. These data are then calibrated to a simulation of the sky containing 2420 point--like objects selected from the MWA Commissioning Survey \citep[MWACS;][]{hur14}. It has a flux density limit of 25~mJy and a declination range of -12\arcdeg~ to -40\arcdeg~ evenly covering the field of view of the observations reported here. The objects used in calibration are selected to lie inside the 5\% contour of the primary lobe of the tile power--pattern. The calibration algorithm -- based on forward modeling software by \citet{sul12} and the calibration method described by \citet{sal14} -- computes complex gain solutions per channel per antenna averaged to two minute intervals. The solutions are fairly low signal--to--noise so we iteratively average along the antenna and frequency dimensions to capture the relatively independent passband and antenna--to--antenna variation. First, we average the channel gains over all antennas to obtain a high signal--to--noise measurement of the bandpass. After applying this single passband, we do a second round of calibration and fit second and first order polynomials for amplitude and phase respectively for each antenna. This flattens any residual variation in bandpass and removes small phase slopes due to variations in cable delay. Finally, we fit for an additional phase known to be caused by small reflections in subset of cables.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=\linewidth]{figures/v1_0/multi_baseline_fhd_delay_spectrum_snapshots.eps}
\caption{Delay spectra of two observations of the EoR field (RA = 0$^\textrm{h}$, Dec = $-30$\arcdeg) at an {\it off--zenith} pointing (top) when tiles have significant secondary lobes in the direction of the Galactic center and a {\it zenith} pointing (bottom) where secondary lobes are minimal and the Galaxy has set. For reference, the $k$--space equivalent of $\tau$ and $|\vec{b}|$, namely, $k_\parallel$ and $k_\perp$ are shown at the axes on the right and top respectively. White lines mark the boundaries of {\it foreground wedge} determined by light travel time across each baseline. Smooth spectrum objects predominantly reside in this {\it foreground wedge}. The logarithmic color scale (shown at the right) is common to both panels. \label{fig:fhd_data}}
\end{figure}

@@

The degree to which these objects leak power into spectral line of sight modes in the {\it EoR window} will be shown to depend primarily on the brightness of foregrounds near the boundaries of the {\it foreground wedge}. Thus, it is important to consider an all--sky model for foreground objects in evaluating the features seen in the power spectrum instead of restricting only to the primary field of view, a point also supported by Pober~et~al.~(2014, in preparation).

@@

@@

A very interesting signature of diffuse emission is revealed at regions near the positive horizon delay limit in the {\it off--zenith} pointing and at both positive and negative horizon delay limits in the {\it zenith} pointing even on wide antenna spacings (125~m~$\lesssim |\vec{b}|\le$~200~m). This is contrary to expectations because these longer baselines are not expected to be sensitive to diffuse emission (on scales $\gtrsim$~0\fdg 85) of our input foreground model. In the wide--field observations of the MWA, long baselines appear shortened due to projection effects for off--axis emission, especially towards the horizon.  This effect is evident at all LSTs in our simulations and it naturally holds for short baselines, as well. Thus, diffuse emission from far off--axis directions manifests as an edge--heavy {\it two--pronged fork} across all baselines. It decreases in strength with increasing baseline length as expected but is present in all baseline orientations.

@@

@@

In principle, instead of discarding selected antenna spacings altogether, we could down--weight them based on an optimal scheme. This technique provides a very simple and yet effective tool in adding a layer of control to mitigate effects of foreground contamination in EoR data analysis.

@@

In conclusion, we find that inclusion of emission models, both diffuse and compact, all the way to the horizon is essential to explaining the observed power spectrum. We also provide a simple and effective tool based on the delay spectrum technique to mitigate foreground contamination by an order of magnitude in EoR data analysis by discarding or down--weighting data from antenna pairs most affected by foreground contamination as a function of antenna spacing vector and time of observation, with negligible loss of sensitivity.
